---
title: "Math 158 Project Outline"
author: "Kat Gelsey, Paul McKinley, Sarah Kim"
date: "11/30/2021"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
## Abstract
A one paragraph summary of what you set out to learn, and what you ended up finding. It

## Introduction
The objective of this project is to analyze how one sector of the U.S. economy (i.e. energy) can influence broader societal outcomes, such as economic growth and elections. In this case, we are interested in building models based on energy supply (for electricity) in each state, to see what role the energy sector may have on a broader outcome.

Questions we would like to address in this paper includes:
(1) Is electricity production type correlated with total GDP per state?
(2) Which sources of electricity production is most significant in predicting GDP?
(3) In addition to energy production, is total energy consumption a significant predictor for total GDP per state?
(4) Is electricity production type correlated with political affiliation 
(via election results) per state?

Our project is broken into three components. First, we analyze total GDP with total energy consumption by state as a predictor to see if total energy consumption is relevant to total GDP.

Then, we aim to investigate the relationship between energy production source and total energy consumption against and total GDP. This allows us to compare both energy production and comsumption in the context of our societal variables of interest. Our energy data comes from the National Energy Institute, providing the percentage of electricity that is produced in a given state by each source, such as solar, wind, coal, natural gas, etc. Our GDP data comes from Statistia Research Department. 

Finally, we are interested in the relationship between energy production and political affiliation. In this case, we use the percentage of voters who voted Republican vs. Democrat in the 2020 presidential race as a response, and energy production type as predictors. 

## Data
Data Set: Describe details about how the data set was collected and the variables in the data set.

**MAKING THE FULL DATASET**
```{r}
require(dplyr)
fulldata <-read.csv("Project Data/masterdata.csv")

#renaming row names into states
rownames(fulldata) = fulldata$STATE[c(1:51)] #changing rownames

```

**response variables: total GDP & Political Affiliation**
```{r}
boxplot(fulldata$total_GDP, xlab = "GDP", main = "Boxplot for Total GDP in year 2020")

#plotting total US GDP
require(tidyverse)
ggplot(data = fulldata, mapping = aes(x = reorder(STATE, total_GDP), y=total_GDP)) + 
  geom_bar(stat = "identity") + coord_flip()+
  labs(x="STATE", y="2020 Total US GDP (in billions of dollars)")
  ggtitle("2020 US GDP (in billion $)")

boxplot(fulldata$dem_percent, xlab = "Republican percentage", main = "")
boxplot(fulldata$rep_percent, xlab = "Democrats percentage", main = "")
```

**explanatory variables ~ total GDP**
```{r}
# response variable GDP ~ some numerical variables of interest
plot(total_GDP ~ COAL, fulldata, main = "Scatterplot for GDP ~ COAL")
abline(lm(total_GDP ~ COAL, fulldata))

plot(total_GDP ~ PETROLEUM, fulldata, main = "Scatterplot for GDP ~ PETROLEUM")
abline(lm(total_GDP ~ PETROLEUM, fulldata))

plot(total_GDP ~ NUCLEAR, fulldata, main = "Scatterplot for GDP ~ NUCLEAR")
abline(lm(total_GDP ~ PETROLEUM, fulldata))

plot(total_GDP ~ WIND, fulldata, main = "Scatterplot for GDP ~ WIND")
abline(lm(total_GDP ~ PETROLEUM, fulldata))
```
**explanatory variables ~ Political Affiliation**
```{r}
# response variable GDP ~ some numerical variables of interest
plot(rep_percent ~ COAL, fulldata, main = "Scatterplot for REPUBLICAN VOTES ~ COAL")
abline(lm(rep_percent ~ COAL, fulldata))

plot(rep_percent ~ PETROLEUM, fulldata, main = "Scatterplot for REPUBLICAN VOTES ~ PETROLEUM")
abline(lm(rep_percent ~ PETROLEUM, fulldata))

plot(rep_percent ~ NUCLEAR, fulldata, main = "Scatterplot for REPUBLICAN VOTES ~ NUCLEAR")
abline(lm(rep_percent ~ PETROLEUM, fulldata))

plot(rep_percent ~ WIND, fulldata, main = "Scatterplot for REPUBLICAN VOTES ~ WIND")
abline(lm(rep_percent ~ PETROLEUM, fulldata))
```

## Analysis
```{r}
library(MASS)
library(faraway)
```

**Model 1: GDP Growth ~ Total Energy Consumption by state**
After we create the model and look at the summary, we see that a state's energy consumption appears to be an extremely significant predictor of the state's GDP. With this predictor alone we get an R$^2$ value of 0.64, and the p-value for the energy consumption predictor is 2.31e-12.
```{r}
GDPcon <- lm(total_GDP ~ consumption_Tbtu, fulldata)
summary(GDPcon)
```
##Boxcox
After creating the model, our first step is to determine whether the response (total GDP) needs to be transformed. The boxcox() function plot tells us that transformation is necessary, since the confidence interval for $\lambda$ is centered at 0.2 and does not contain 1. We should transform the data by raising the response to the 1/5th. 
```{r}
boxcox(GDPcon, lambda=seq(-0.25,0.75,by=0.05),plotit=T)

GDPcon_trans <- lm((total_GDP)^(0.2) ~ consumption_Tbtu, fulldata) 
sumary(GDPcon_trans)
```

##Diagnostics
*Normality*
ORIGINAL DATA: 
Even for the transformed data, we see that the plot of the residuals is not very random, with the largest fitted values having extremely high or low residuals. To check for normality, we look at a QQ plot of the residuals. There is some short-tailed distribution, but not as bad as when I plotted the QQ plot with the untransformed data, which had a long-tailed distribution. The Shapiro-Wilk test demonstrates that the data is now normal, versus the test's highly significant p-value using the untransformed data. (Faraway 75-81)
```{r}
plot(GDPcon_trans)
shapiro.test(residuals(GDPcon_trans))
```

Furthermore, it appears that 33 (New York) and 44 (Texas) are leverage points, influential points, and outliers.

Removing Texas from the model significantly improves the R$^2$ value for the model and drastically increases the significance of the consumption_Tbtu predictor. Therefore, Texas is an outlier and an influential point.
```{r}
fulldata[33,]#new york
GDPcon_trans_33 <- lm((total_GDP)^(0.2) ~ consumption_Tbtu, fulldata[-33,])
sumary(GDPcon_trans_33)

fulldata[44,]#texas
GDPcon_trans_44 <- lm((total_GDP)^(0.2) ~ consumption_Tbtu, fulldata[-44,])
sumary(GDPcon_trans_44)
```
(Since we have only one predictor, it is irrelevant to consider collinearity, stepwise/backwards regression, or PCR.)

**Model 2: total GDP ~ Energy production source by type for each state**
Our second model is to look at how energy production by type could be a predictor of gross domestic product. First, we want to see if we need to transform our response variable using the boxcox(). Boxcox() plot suggests that we transform our response variable by taking its log. 
```{r}
GDPpro <- lm(total_GDP ~ NUCLEAR + COAL+ NATURAL.GAS + PETROLEUM + HYDRO + GEOTHERMAL + SOLARPV + WIND + BIOMASS_OTHER, fulldata)

# boxcox 
bc <- boxcox(GDPpro, lambda=seq(-0.5,0.5,by=0.05), plotit = T) 
```
With our second model, we want to find which energy production by type are significant predictors for state GDP. Setting GDP as response and energy production types as predictors, we constructed a model with all energy type. Then, we ran a step regression model. Just looking at the energy production types as predictors. We find that nuclear, natural gas, and solar energy are significant predictors of GDP
```{r}
# linear model with renewable energy source by production type and total energy consumption as predictors
GDPpro_log <- lm(log(total_GDP) ~ NUCLEAR + COAL+ NATURAL.GAS + PETROLEUM + HYDRO + GEOTHERMAL + SOLARPV + WIND + BIOMASS_OTHER, fulldata)
summary(GDPpro_log)

#step()
step(GDPpro_log)
GDPpro_log_step <- lm(formula = log(total_GDP) ~ NUCLEAR + COAL + NATURAL.GAS + 
    PETROLEUM + HYDRO + SOLARPV + WIND, data = fulldata)
summary(GDPpro_log_step)
```
Moving forward, we wanted to construct a model that predicts GDP with both total energy consumption and energy production source types as predictors. We then examined, if energy production by type is just enough of variables to predict GDP by running an ANOVA analysis.
```{r}
GDP <-lm(total_GDP ~ NUCLEAR + COAL+ NATURAL.GAS + PETROLEUM + HYDRO + GEOTHERMAL + SOLARPV + WIND + BIOMASS_OTHER + consumption_Tbtu, fulldata)
bc <- boxcox(GDP, lambda=seq(-0.5,0.5,by=0.05), plotit = T) 

GDPfull <-lm(log(total_GDP) ~ NUCLEAR + COAL+ NATURAL.GAS + PETROLEUM + HYDRO + GEOTHERMAL + SOLARPV + WIND + BIOMASS_OTHER + consumption_Tbtu, fulldata)
anova(GDPfull, GDPpro_log)
```
Since the P-value for the ANOVA diagnostics are smaller than 0.05. We can conlude that the reduced model does not capture as much as the larger model. Hence, the bigger model (with both consumption and production as predictors are preferred). Hence, we decided to continue with the bigger model 'GDPfull' and run stepwise regression analysis to see which explanatory variables are significant.
```{r}
# step(GDPfull)
GDPfull_step <- lm(formula = log(total_GDP) ~ NUCLEAR + SOLARPV + consumption_Tbtu, 
    data = fulldata)
summary(GDPfull_step)
```
Our stepwise regression model concluded that nuclear energy production, solar energy production, and total energy consumption are significant predcitros. 
```{r}
#anova
anova(GDPfull, GDPfull_step)#reduced is fine
```
Comparing our models--one with all predictors and the reduced model--on ANOVA, we can conclude that the reduced model does as well as the bigger model. We can conclude that nuclear energy, solar energy, and total energy consumptions are indeed significant predictors. 

Now we want to see the confidence interval for the model
```{r}
# 95% CI
confint(GDPfull_step)#all significant
```


##Model Diagnostics
We can take a look at the distribution of residuals and leverage points in model:
```{r}
plot(GDPfull_step)
```
From the model, we can see there is an underlying structure for residuals and fitted model. The model, therefore, does not satisfy constant variance. 

##Check for Normality
```{r}
shapiro.test(residuals(GDPfull_step))
```
Based on the shapiro-wilk normality test, the residuals are normally distributed either.

## Large leverage points.
```{r}
hatv1 <- hatvalues(GDPfull_step)
p <- sum(hatv1)
n <- 51
2*p/n

rev(sort(hatv1))
fulldata[44,]
fulldata[5,]
fulldata[29,]
fulldata[30,]
```
Texas, California, Nevada, New Hampshire significant leverage points because they exceed 2p/n = 0.157

## Outliers
```{r}
stud <- rstudent(GDPfull_step)
stud[which.max(abs(stud))]

qt(0.05/(n*2),n-p)

rev(sort(abs(stud)))
```
Based on the Bonferroni critical value, Texas is an outlier, 

## Influential Points
```{r}
cook <- cooks.distance(GDPfull_step)
halfnorm(cook, ylab="Cook's Distances")

fulldata[44,] #Texas
fulldata[30,] #New Hampshire
```
Texas and New Hampshire are influential points.

## Analysis on Outliers, Leverage Points, Influential Points
```{r}
#plotting NUCLEAR
require(tidyverse)
ggplot(data = fulldata, mapping = aes(x = reorder(STATE, NUCLEAR), y=NUCLEAR)) + 
  geom_bar(stat = "identity") + coord_flip()+
  labs(x="STATE", y="Nuclear Energy Production (%)")
  ggtitle("NUCLEAR")

#plotting SOLAR
require(tidyverse)
ggplot(data = fulldata, mapping = aes(x = reorder(STATE, SOLARPV), y=SOLARPV)) + 
  geom_bar(stat = "identity") + coord_flip()+
  labs(x="STATE", y="SOLAR Energy Production (%)")
  ggtitle("SOLAR")
  
#plotting Consumption
require(tidyverse)
ggplot(data = fulldata, mapping = aes(x = reorder(STATE, consumption_Tbtu), y=consumption_Tbtu)) + 
  geom_bar(stat = "identity") + coord_flip()+
  labs(x="STATE", y="Total Energy Consumption")
  ggtitle("Total Energy Consumption ($)")


```

Plotting the graphs for the significant variables, indeed it mathces with outliers, influential points, and leverage points

-> look at the consumption data for these states and see what they have to say. 
-> inhibits our predictive powers in our model, however, low r squared makes sense because we need lots of other variables to measure GDP. lots of other varialbes (not high r square) 

## Prediction
```{r}
# Alabama as a test state
GDPreduc_Al <-lm(log(total_GDP) ~ NUCLEAR + SOLARPV + consumption_Tbtu, fulldata[-1,]) #take out WA #take out a value that's not too extreme

testAlabama <-fulldata[1,] %>%
  dplyr::select(NUCLEAR, SOLARPV, consumption_Tbtu)

exp(predict(GDPreduc_Al,testAlabama, interval = 'prediction'))
fulldata[1,]$total_GDP

# Washington as a test state
GDPreduc_WA <-lm(log(total_GDP) ~ NUCLEAR + SOLARPV + consumption_Tbtu, fulldata[-48,]) #take out WA #take out a value that's not too extreme

testWash <-fulldata[48,] %>%
  dplyr::select(NUCLEAR, SOLARPV, consumption_Tbtu)

exp(predict(GDPreduc_WA,testWash, interval = 'prediction'))
fulldata[48,]$total_GDP

```

**Model 3: Political Affiliations ~ Energy production source by type for each state**
We are also interested in energy production by type (i.e. Solar, wind, petroleum) as predictors for political affiliation. In this case, we can construct models with our breakdown of energy types by state as predictors and the percentage of voters who voted Republican and who voted Democrat. 

A model with percentage of democratic voters as a response and energy types as predictors. 
```{r}
demMdl<-lm(dem_percent~NUCLEAR+COAL+NATURAL.GAS+PETROLEUM+HYDRO+GEOTHERMAL+SOLARPV+WIND+BIOMASS_OTHER,data = fulldata)
summary(demMdl)
```

A similar comparison with Republican voter percentage:
```{r}
repMdl<-lm(rep_percent~NUCLEAR+COAL+NATURAL.GAS+PETROLEUM+HYDRO+GEOTHERMAL+SOLARPV+WIND+BIOMASS_OTHER,data = fulldata)
summary(repMdl)
```

Since we are using the percentage of votes as our response, our models for the Democrat vs. Republican percentages should yield very similar results, in terms of significant predictors. For now, let's focus on Republican response data. Before conducting additional analysis, let us consider the addition of other economic variables, specifically Total GDP and Total Energy Consumption. 

```{r}
#updating the model
demMdl2<-update(demMdl,.~.+total_GDP +consumption_Tbtu)
repMdl2<-update(repMdl,.~.+total_GDP +consumption_Tbtu)

#summarize both updated models
summary(demMdl2)
summary(repMdl2)
```
We see that total_GDP is a significant predictor and our R-squared value has increased, as might be expected with the addition of new predictors. However, we still find that most of our predictors are not very significant. To determine if we should, in fact, include these new predictors in the updated model, we can run a analysis of variance on both the models. 

```{r}
# republican
anova(repMdl,repMdl2)

# democrat
anova(demMdl,demMdl2)
```
Our null hypothesis for the ANOVA comparison is that there is not significant difference in the predictive power of the small model and the large model that includes the economic predictors. Our p-value for both comparisons is greater than 0.1, disallowing us from rejecting the null hypothesis. Therefore, we can use our smaller models repMdl', which include only energy production by type as predictors. This also focuses our analysis on energy type.

##Transformations
We want to determine if a transformation of our response data is needed. We can use a Box-Cox analysis to do this:
```{r}
require(MASS)
boxcox(repMdl,plotit=T,lambda = seq(-1,4))
```
Our peak is centered around 0.3, although 1 is still contained in our confidence interval, suggesting a transformation is not likely necessary. Here, we do not apply a transformation of the response. 

##Stepwise regression
It is unclear if all of our predictors are necessary to construct a useful model. We use a stepwise regression to evaluate which predictors are most impactful. The stepwise regression tells us that SOLAR and BIOMASS could be removed from the model, which is a surprising deviation from our first model with GDP as the response. We can construct an updated version of repMdl with these predictors removed.
```{r}
step(repMdl)
stepRepMdl<-lm(formula = rep_percent ~ NUCLEAR + COAL + NATURAL.GAS + PETROLEUM + 
    HYDRO + GEOTHERMAL + WIND, data = fulldata)
summary(stepRepMdl)
```

Removing SOLARPV and BIOMASS_OTHER shows that our R-squared value has increased, and we now see all of our predictors are significant at the 5% level or better. 

## Model Diagonostic
```{r}
plot(stepRepMdl)
```
##Predicting with our model on one state's data
```{r}
require(tidyverse)
n=10
repMdlPredict <-lm(rep_percent~NUCLEAR+COAL+NATURAL.GAS+PETROLEUM+HYDRO+GEOTHERMAL+WIND, fulldata[-n,]) #take out Alabama #take out a value that's not too extreme
summary(repMdlPredict)
testAlabamaPolit <-fulldata[n,]
testAlabamaPolit<- testAlabamaPolit%>%
  dplyr::select(NUCLEAR,COAL,NATURAL.GAS,PETROLEUM,HYDRO,GEOTHERMAL,WIND)
#testAlabama <-data.frame(NUCLEAR = 32.0, SOLARPV = 0.3, consumption_Tbtu = 1954.8)
predict(repMdlPredict,testAlabamaPolit,interval = "prediction")
fulldata[n,]$rep_percent
```

##Limitation: Correlation between variables
Let us continue working with the model with Republican voter percentage as the response. We are interested in the potential correlations amongs predictors. 
We can construct a model matrix, and compute correlations between predictors. 
```{r}
require(faraway)
repMat<-model.matrix(stepRepMdl)
cor(repMat)
```
There is a somewhat surprising level of correlation between solar and coal, as well as natural gas and geothermal; however, for the most part our predictors appear uncorrelated. 

Checking for evidence of collinearity
```{r}
require(faraway)
vif(stepRepMdl)
```

## Principal Components Regression
We are also interested in reducing dimensionality and effects of collinearity in our model. We can perform a principal components regression (PCR) which makes use of principal component analysis (PCA) to construct a model based on "components" that explain some amount of variability in the data, comprised of linear combinations of our predictors. 
```{r}
# Constructing a PCR model:
require(pls)
set.seed (123)
pcr_model <- pcr(rep_percent~NUCLEAR+COAL+NATURAL.GAS+PETROLEUM+HYDRO+GEOTHERMAL+SOLARPV+WIND+BIOMASS_OTHER, data = fulldata, scale = TRUE, validation ="CV")

summary(pcr_model)
```
Plotting our results: 
```{r}
validationplot(pcr_model,val.type = "R2")
```

